Quantum H theorem, variational entropy, and relaxation processes

### 2020-10-16 notasparaarticulo.tex
discussions and remarks section was included, conclusions was rewritten, introduction was modified and some new ideas were included

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{xcolor}
\usepackage{calrsfs}
\DeclareMathAlphabet{\pazocal}{OMS}{zplm}{m}{n}

@@

\textcolor{red}{In the literature, the theoretical basis to describe equilibrium systems in the classical scheme is well-established. From thermodynamics to statistical mechanics, the procedures to describe mathematically equilibrium systems for both schemes (phenomenological for thermodynamics and microscopical for statistical mechanics) are well-known and works in almost every system in nature \cite{huang, reif, patrick}. In the case of the out of equilibrium systems, there are many works that attempt to describe out of equilibrium systems (however those works do not work in some systems, for example, the internal behavior of a glass[], which behavior is too slow compare with the time observation). Examples of some theoretical schemes to describe out of equilibrium systems can be the Onsager formulation (called linear thermodynamics), some few other restricted developments that attempt to complete the Onsager formulation \cite{kei, onsager} and the Kinetic Theory of dilute gases developed by Boltzmann.\\

@@

In this work we define a new $H$ functional (variational entropy) which depends on the distribution function of an ideal quantum gas (Bosons or Fermions particles). This gas is contained inside a volume $V$ divided into cells. Before developing the non-equilibrium description, we recover the Fermi-Dirac and Bose-Einstein distributions from the variational entropy when the system is in the equilibrium state. Furthermore we prove that the derivative of the variational entropy with respect to time is less or equal to zero for relaxation processes (the action of any perturbed system that returns to the equilibrium state) in any system. This prove corresponds with the statement of the $H$ theorem. In addition we obtain the Boltzmann transport equation from the Boltzmann's $H$ functional and, in analogy with this, we find a quantum version of the Boltzmann transport equation from the variational entropy. Finally, in order to be consistence with the correspondence principle, we prove that the Boltzmann's $H$ functional is proportional to the variational entropy in the limit of the high level energies.}

@@

\textcolor{red}{We begin by defining the form of the functional $H$

@@

    H(t)&=&\sum_{M} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\

@@

local identical volumes (\textit{i.e.} $V_M = V/K, \forall M$). Each cell has its own local chemical potential, $\mu_M$, local
$H$ functional, $H_M$, local number of particles, $\mathcal{N}_M$, and local energy, $\mathcal{E}_M$. In terms of the local distribution

@@

    H_M&=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \delta \epsilon_n.

@@

    &&\sum_{M} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\right]\delta V_M \delta \epsilon_n=\sum_{M} {\mathcal{N}}_{M} \delta \epsilon_n=N; \nonumber \\
    &&\sum_{M}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{Mn}\right]\delta V_M \delta g_n=\sum_M {\mathcal{E}}_M \delta \epsilon_n=E, \label{restriccions1}

@@

Inspired by the Hamilton principle, $H$ will be maximized with those restrictions to obtain the equilibrium distribution functions for quantum gasses using the Lagrange multipliers method obtaining

@@

      H(t)&=&\sum_M \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right]\delta V_M \delta \epsilon_n\label{H-entropy}.

@@

\begin{equation}
   H'(t)=\int_{}^{}\int_{}^{} f(r,v,t) \ln f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r \label{CH2}.
\end{equation}
The functional (\ref{CH2}) contains a no-homogeneous distribution function but keeps the same form of the Boltzman's $H$ functional. Excluding the homogeneous distribution function hypothesis, we use the same procedure that it was shown in the previous section to prove the $H$-theorem.\\ 
If one considers the microcanonical ensemble conditions 
\begin{equation}
    \int_{}^{}\int_{}^{}f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r=N, \ \ \ \int_{}^{}\int_{}^{}f(r,v,t)\epsilon(v)\mathrm{d}^3v \mathrm{d}^3r=E \label{micro},
\end{equation}{}
one obtains the distribution function that minimizes $H$
\begin{eqnarray}
\frac{\delta H'}{\delta f(r',v')}&=&\int\int_{}^{}\frac{\delta}{\delta f(r',v')}\left[f(r,v)\ln f(r,v)  \right]\mathrm{d}^3v \mathrm{d}^3r-\int_{}^{}\alpha(r)\int_{}^{}\frac{\delta f(r,v)}{\delta f(r',v')}\mathrm{d}^3v \mathrm{d}^3r\nonumber \\
&&-\int_{}^{}\beta(r) \int_{}^{}\epsilon(v)\frac{\delta f(r,v)}{\delta f(r',v')}\mathrm{d}^3v \mathrm{d}^3r\nonumber \\
&=&\ln f(r',v')+1-\alpha(r')-\beta(r') \epsilon(v')=0.
\end{eqnarray}{}
isolating the distribution function 
\begin{eqnarray}
\ln f(r',v')&=&\alpha(r')+\beta(r') \epsilon(v')-1 \ \ \  \Rightarrow \ \ \ f(r',v')=e^{\alpha(r') +\beta(r') \epsilon(v')-1} \nonumber \\
&=&Ce^{\alpha(r')+\beta(r') \epsilon(v') } \label{relacion1},
\end{eqnarray}{}
where $C$ is a constant. It is trivial that using the variational procedures, one obtains the form of the Maxwell-Boltzmann distribution. \\
As $H'$ in equilibrium \textcolor{red}{is proportional to} entropy, it is trivial to think that Lagrange multipliers do not depend on the position because of the system is in equilibrium and consequently, the distribution function is homogeneous, that is
\begin{equation}
    f(v)=Ce^{\alpha+\beta \epsilon(v)}.
\end{equation}{} 
The constant $C$ can be omitted defining the following $H$ functional
\begin{equation}
   H''(t)=\int_{}^{}\int_{}^{} \left[f(r,v,t) \ln f(r,v,t)-f(r,v,t)\right]\mathrm{d}^3v \mathrm{d}^3r \label{CH3},
\end{equation}{}
  but $\int \int f(r,v,t)\mathrm{d}^3v \mathrm{d}^3r=N$ where $N$ is the total particle number. As $N$ is a constant and any $H$ functional has sense when we calculate its time derivative, this constant can be omitted. 
\section{$H$-functional for relaxation processes in out of equilibrium systems}

@@

The \textit{variational entropy} has the following form
\begin{equation}
    H(t)=\sum_{M} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right]\delta V_M \delta \epsilon_n \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.
This distribution function \textcolor{red}{must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells)}, and the following microcanonical restrictions
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. Those condition described a system out of equilibrium but no so far from it, that is a perturbed system.  \\
If we perform the derivative of the functional $H$ with respect to time, we obtain
\begin{equation}
   \frac{dH(t)}{dt}= \sum_n \sum_M \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right]\delta V_M \delta \epsilon_n.\label{deltaH}
\end{equation}{}
Using the first-order approximation
\begin{eqnarray}
   f_{Mn}=\bar{f}_{n}(1+g_{Mn}): \ \ \ \bar{f}_{n}\gg \bar{f}_{n}|g_{Mn}|, \label{firstorder}
\end{eqnarray}{}
we get

@@

    \frac{dH(t)}{dt}&=&\sum_n \sum_M \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM}\delta V_M \delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_M \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\delta V_M \delta \epsilon_n.\nonumber \\

@@

    \frac{dH}{dt}&=&\sum_n \sum_M \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\}\delta V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\delta V_M \delta \epsilon_n.\label{cambioH2}

@@

    \frac{dH}{dt}&=&\sum_n \sum_M \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \}\delta V_M \delta \epsilon_n. \nonumber \\

@@

and as a consequence of $\sum_M \Delta_M(t) \delta V_M =\sum_M \delta_M(t)\delta V_M =0$, we find

@@

    \sum_M \dot{\Delta}_M(t)\delta V_M =\sum_M \dot{\delta}_{M}(t)\delta V_M =0.

@@

   \frac{dH}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M  g_{nM}\dot{g}_{nM}\delta V_M \delta \epsilon_n. \nonumber \\ \label{cambioH4}

@@

    \frac{dH}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\delta V_M \delta \epsilon_n+\sum^{N}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ}\delta V_M \delta \epsilon_n \right), \label{cambioH5}

@@

    \frac{dH}{dt}=-\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} |g^{+}_{nJ}||\dot{g}^{+}_{nJ}|\delta V_M \delta \epsilon_n+\sum^{N}_J  |g^{-}_{nJ}||\dot{g}^{-}_{nJ}|\delta V_M \delta \epsilon_n \right), \label{cambioH6}

@@

and due to $e^{\bar{\alpha}+\bar{\beta}\epsilon_n}$ is always positive, then $\frac{dH}{dt}<0$. With this, we proved that any quantum ideal gas perturbed (out of equilibrium but not so far from it), evolves such that $\frac{dH}{dt}<0$.\\
When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{dH}{dt}=0$. Then, we can say that any quantum ideal gas perturbed, evolves always to the equilibrium state, such as the classical $H$ theorem. \\

@@

\subsection{Clasical scheme}
Now we use the restrictions (\ref{restrictionoutside}) but in the form of the continuous case
\begin{equation}
    \int_{}^{}f(r,v,t)\mathrm{d}^3v=\mathcal{N}+\Delta(r,t); \ \ \ \int_{}^{}f(r,v,t)\epsilon(v)\mathrm{d}^3v=\mathcal{E}+\Omega(r,t)\label{restriccionescambio},
\end{equation}{}
where $\Delta(r,t)$ and $\Omega(r,t)$ could be seen as a deviation from the mean number of particles and mean energy by differential element $d^3r$.
Performing the derivative of the functional $H$ (\ref{CH2}) with respect to time
\begin{equation}
    \frac{dH'}{dt}=\int_{}^{}\int_{}^{}\left[ 1+\ln f(r,v,t) \right]\dot f(r,v,t) \mathrm{d}^3v \mathrm{d}^3r \label{dH1},
\end{equation}{}
and applying the first order approximation (\ref{firstorder}) to (\ref{dH1}), we obtain
\begin{equation}
    \frac{dH'}{dt}=\int_{}^{}\int_{}^{}\bar f(v) \left [ 1+\ln \left\{ \bar f(v)+\bar f(v)g(r,v,t) \right\} \right]\dot g(r,v,t)\mathrm{d}^3v \mathrm{d}^3r \label{dH1,1}.
\end{equation}{}
Using the same approximation (\ref{lnapproximation})
to (\ref{dH1,1}) 
\begin{eqnarray}
\frac{dH'}{dt}&=&\int_{}^{} \int_{}^{} \bar f(v)\left[ 1+\ln \bar f(v)+g(r,v,t) \right]\dot g(r,v,t)\mathrm{d}^3v \mathrm{d}^3r,
\end{eqnarray}{}
using (\ref{relacion1}) with homogeneous multipliers
\begin{eqnarray}
\frac{dH'}{dt}&=&\int_{}^{}\int_{}^{}\bar f(v)\left[ \alpha+\beta \epsilon(v) \right]\dot g(r,v,t)\mathrm{d}^3v \mathrm{d}^3r+\int_{}^{}\int_{}^{}\bar f(v)g(r,v,t)\dot g(r,v,t)\mathrm{d}^3v \mathrm{d}^3r \label{dH1,2}. \nonumber \\
\end{eqnarray}{}
If we use the same procedure shown in the previous section to the restrictions (\ref{restriccionescambio}), we obtain the same result. Then applying this to (\ref{dH1,2}), we find
\begin{eqnarray}
\frac{dH'}{dt}&=&\int_{}^{}\int_{}^{}\bar f(v)g(r,v,t)\dot g(r,v,t)\mathrm{d}^3v \mathrm{d}^3r.
\end{eqnarray}{}
Using the same consideration in the previous section to $g$ and $\dot g$, that is to divide the spatial integral in two, one integral for cells that have excess of particles and/or energy, and the second for cells that have missing of particles and/or energy, and then using (\ref{separacion}), we prove that $\frac{dH'}{dt}<0$. If the system is in equilibrium, $g(r,v,t)=\dot g(r,v,t)=0$, therefore $\frac{dH'}{dt}=0$. 
\\
\\
The next step in this model is to obtain the analogous transport equation through variation methods. In order to verify that transport equation can be obtained by applying the variation method, in the next section the Boltzmann transport equation will be obtained applying the mentioned method to the modified Boltzmann $H$ functional (\ref{CH2}).

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{xcolor}
\usepackage{calrsfs}
\DeclareMathAlphabet{\pazocal}{OMS}{zplm}{m}{n}
\newcommand{\Sm}{\Ss_M}
\newcommand{\de}{\delta}
\newcommand{\ep}{\epsilon}
\newcommand{\Ss}{\mathcal{S}}

@@

\textcolor{red}{In the literature, the theoretical basis to describe equilibrium systems in the classical scheme is well-established. From thermodynamics to statistical mechanics, the procedures to describe mathematically equilibrium systems for both schemes (phenomenological for thermodynamics and microscopical for statistical mechanics) are well-known and works in almost every system in nature \cite{huang, reif, patrick}. In the case of the out of equilibrium systems, there are many works that attempt to describe out of equilibrium systems (however those works do not work in some systems, for example, the internal behavior of a glass due to the time relaxation is very big \cite{cristal1} and the entropy measurement \cite{cristal2, cristal3} as examples). Examples of some theoretical schemes to describe out of equilibrium systems can be the Onsager formulation (called linear thermodynamics), some few other restricted developments that attempt to complete the Onsager formulation \cite{kei, onsager} and the Kinetic Theory of dilute gases developed by Boltzmann.\\

@@

In this work we define a new functional (variational entropy) which depends on the distribution function of an ideal quantum gas (Bosons or Fermions particles). This gas is contained inside a volume $V$ divided into cells. Before developing the non-equilibrium description, we recover the Fermi-Dirac and Bose-Einstein distributions from the variational entropy when the system is in the equilibrium state. Furthermore we prove that the derivative of the variational entropy with respect to time is less or equal to zero for relaxation processes (the action of any perturbed system that returns to the equilibrium state) in any system. This prove corresponds with the statement of the $H$ theorem. In addition we obtain the Boltzmann transport equation from the Boltzmann's $H$ functional and, in analogy with this, we find a quantum version of the Boltzmann transport equation from the variational entropy. Finally, in order to be consistence with the correspondence principle, we prove that the Boltzmann's $H$ functional is proportional to the variational entropy in the limit of the high level energies.}

@@

\textcolor{red}{We begin by defining the form of the functional 

@@

    \Ss (t)&=&\sum_{M} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\

@@

local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each cell has its own local chemical potential, $\mu_M$, local
variational functional, $\Sm$, local number of particles, $\mathcal{N}_M$, and local energy, $\mathcal{E}_M$. In terms of the local distribution 

@@

    \Sm &=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \de V_M \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \de V_M \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \de V_M \delta \epsilon_n.

@@

    &&\sum_{M} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\de \epsilon_n\right] \de V_M=\sum_{M} {\mathcal{N}}_{M} \de V_M=N; \nonumber \\
    &&\sum_{M}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\delta \epsilon_n\right] \de V_M=\sum_M {\mathcal{E}}_M \delta V_M=E, \label{restriccions1}

@@

Inspired by the Hamilton principle, $\Ss$ will be maximized with those restrictions to obtain the equilibrium distribution functions for quantum gasses using the Lagrange multipliers method obtaining

@@

      \Ss(t)&=&\sum_M \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right] \de V_M \delta \epsilon_n\label{H-entropy}.

@@

\begin{equation}
   H'(t)=\sum_M\int_{}^{} f_M(v,t) \ln f_M(v,t)\mathrm{d}^3v  \label{CH2}.
\end{equation}
The functional (\ref{CH2}) contains a no-homogeneous distribution function but keeps the same form of the Boltzman's $H$ functional. Excluding the homogeneous distribution function hypothesis, we use the same procedure that it was shown in the previous section to prove the $H$-theorem.\\ 
If one considers the microcanonical ensemble conditions 
\begin{equation}
    \sum_M\int_{}^{}f_M(v,t)\mathrm{d}^3v =N, \ \ \ \sum_M\int_{}^{}f_M(v,t)\epsilon(v)\mathrm{d}^3v =E \label{micro},
\end{equation}{}
one obtains the distribution function that minimizes $H$
\begin{eqnarray}
\frac{\delta H'}{\delta f_J(v')}&=&\sum_M\int_{}^{}\frac{\delta}{\delta f_J(v')}\left[f_M(v)\ln f_M(v)  \right]\mathrm{d}^3v -\sum_M\alpha_M\int_{}^{}\frac{\delta f_M(v)}{\delta f_J(v')}\mathrm{d}^3v \nonumber \\
&&-\sum_M\beta_M \int_{}^{}\epsilon(v)\frac{\delta f_M(v)}{\delta f_J(v')}\mathrm{d}^3v \nonumber \\
&=&\ln f_J(v')+1-\alpha_J-\beta_J \epsilon(v')=0.
\end{eqnarray}{}
isolating the distribution function 
\begin{eqnarray}
\ln f(r',v')&=&\alpha(r')+\beta(r') \epsilon(v')-1 \ \ \  \Rightarrow \ \ \ f_J(v')=e^{\alpha_J +\beta_J \epsilon(v')-1} \nonumber \\
&=&Ce^{\alpha_J+\beta_J \epsilon(v') } \label{relacion1},
\end{eqnarray}{}
where $C$ is a constant. It is trivial that using the variational procedures, one obtains the form of the Maxwell-Boltzmann distribution. \\
As $H'$ in equilibrium \textcolor{red}{is proportional to} entropy, it is trivial to think that Lagrange multipliers do not depend on the position because of the system is in equilibrium and consequently, the distribution function is homogeneous, that is
\begin{equation}
    f(v)=Ce^{\alpha+\beta \epsilon(v)}.
\end{equation}{} 
The constant $C$ can be omitted defining the following $H$ functional
\begin{equation}
   H''(t)=\sum_M\int_{}^{} \left[f_M(v,t) \ln f_M(v,t)-f_M(v,t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}{}
  but $\sum_M \int f_M(v,t)\mathrm{d}^3v =N$ where $N$ is the total particle number. As $N$ is a constant and any $H$ functional has sense when we calculate its time derivative, this constant can be omitted. \\
  \textcolor{green}{In other hand, if the distribution function is spatial-homogeneous, $f$ does not depend on the cell number $M$, then we rewrite (\ref{CH2}) as
  \begin{equation}
      H'(t)=\int \sum_M [f(v,t)\ln f(v,t)] \mathrm{d}^{3}v = P\int  f(v,t) \ln f(v,t) \mathrm{d}^{3}v= P H_{boltz}(t), 
  \end{equation}
  where
  \begin{equation}
      H_{boltz}(t)=\int  f(v,t) \ln f(v,t) \mathrm{d}^{3}v.
  \end{equation}
  Besides, if we analyse the following expression
  \begin{eqnarray}
      \int Pf(v,t) \ln [Pf(v,t)] \mathrm{d}^{3}v = \int [(P\ln P)f(v,t) + Pf(v,t) \ln f(v,t)]\mathrm{d}^{3}v, \label{sum-h}
  \end{eqnarray}
  we can observe if the number of particles in the $\mu$-space $f(v,t)$ is to big compare with the number of cells $P$, then the the first term in (\ref{sum-h}) is negible and and consequently we conclude that
  \begin{equation}
      \int f'(v,t) \ln f'(v,t) \mathrm{d}^{3}v = P\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v; \ \ \ f'(v,t)= Pf(v,t). \label{aditive-property} 
  \end{equation}
  We obtained the additive property when we supposed the distribution function is spatial-homogeneous.}

%----------------------------------------------

\section{Variational entropy for relaxation processes in out of equilibrium systems}

@@

The \textit{variational entropy} has the following form
\begin{equation}
    \Ss(t)=\sum_{M} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.
This distribution function \textcolor{red}{must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells)}, and the following microcanonical restrictions
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. Those condition described a system out of equilibrium but no so far from it, that is a perturbed system.  \\
If we perform the derivative of the variational entropy with respect to time, we obtain
\begin{equation}
   \frac{d \Ss (t)}{dt}= \sum_n \sum_M \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}{}
Using the first-order approximation
\begin{eqnarray}
   f_{Mn}=\bar{f}_{n}(1+g_{Mn}): \ \ \ \bar{f}_{n}\gg \bar{f}_{n}|g_{Mn}|, \label{firstorder}
\end{eqnarray}{}
we get

@@

    \frac{d\Ss (t)}{dt}&=&\sum_n \sum_M \bar{f}_{n}\ln \left[ \frac{\bar{f}_{n}(1+g_{nM})}{1\mp \bar{f}_{n} (1+ g_{nM})} \right]\dot{g}_{nM} \de V_M\delta \epsilon_n \nonumber \\
    &=&\sum_n \sum_M \bar{f}_n \left \{ \ln [\bar{f}_n+\bar{f}_n g_{nM}]\dot{g}_{nM}-\ln [1\mp\bar{f}_n\mp\bar{f}_n g_{nM}]\dot{g}_{nM}  \right \}\de V_M \delta \epsilon_n.\nonumber \\

@@

    \frac{d\Ss}{dt}&=&\sum_n \sum_M \bar{f}_n\left \{ (\ln \bar{f}_n+ g_{nM})\dot{g}_{nM}\right\} \de V_M \delta \epsilon_n \nonumber \\
    &&-\sum_{n}\sum_{M}\bar f_n\left\{ \left( \ln[1\mp\bar{f}_n]\mp \left[\frac{\bar{f}_n}{1\mp\bar{f}_n} \right] g_{nM}\right)\dot{g}_{nM} \right \}\de V_M \delta \epsilon_n.\label{cambioH2}

@@

    \frac{d\Ss}{dt}&=&\sum_n \sum_M \bar{f}_n\left \{ (\bar{\alpha}+\bar{\beta}{\epsilon}_n)\dot{g}_{nM}+ g_{nM}\left(1\pm e^{\bar{\alpha}+\bar{\beta}{\epsilon}_n}\right)\dot{g}_{nM} \right \} \de V_M \delta \epsilon_n. \nonumber \\

@@

and as a consequence of $\sum_M \Delta_M(t) \de V_M  =\sum_M \delta_M(t) \de V_M =0$, we find

@@

    \sum_M \dot{\Delta}_M(t) \de V_M =\sum_M \dot{\delta}_{M}(t) \de V_M=0.

@@

   \frac{d\Ss}{dt}&=&  \sum_n e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\sum_M  g_{nM}\dot{g}_{nM} \de V_M\delta \epsilon_n. \nonumber \\ \label{cambioH4}

@@

    \frac{d\Ss}{dt}=\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} g^{+}_{nJ}\dot{g}^{+}_{nJ}\de V_M \delta \epsilon_n+\sum^{N}_J  g^{-}_{nJ}\dot{g}^{-}_{nJ} \de V_M\delta \epsilon_n \right), \label{cambioH5}

@@

    \frac{d\Ss}{dt}=-\sum_n  e^{\bar{\alpha}+\bar{\beta}\epsilon_n}\left(\sum_J ^{L} |g^{+}_{nJ}||\dot{g}^{+}_{nJ}|\de V_M \delta \epsilon_n+\sum^{N}_J  |g^{-}_{nJ}||\dot{g}^{-}_{nJ}| \de V_M \delta \epsilon_n \right), \label{cambioH6}

@@

and due to $e^{\bar{\alpha}+\bar{\beta}\epsilon_n}$ is always positive, then $\frac{d\Ss}{dt}<0$. With this, we proved that any quantum ideal gas perturbed (out of equilibrium but not so far from it), evolves such that $\frac{dH}{dt}<0$.\\
When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can say that any quantum ideal gas perturbed, evolves always to the equilibrium state, such as the classical $H$ theorem. \\

@@

\subsection{Clasical scheme}
Now we use the restrictions (\ref{restrictionoutside}) but in the form of the continuous case
\begin{equation}
    \int_{}^{}f_M(v,t)\mathrm{d}^3v=\mathcal{N}+\Delta_M(t); \ \ \ \int_{}^{}f_M(v,t)\epsilon(v)\mathrm{d}^3v=\mathcal{E}+\Omega_M(t)\label{restriccionescambio},
\end{equation}{}
where $\Delta_M(t)$ and $\Omega_M(t)$ could be seen as a deviation from the mean number of particles and mean energy.

Performing the derivative of the functional $H$ (\ref{CH2}) with respect to time
\begin{equation}
    \frac{dH'}{dt}=\sum_M\int_{}^{}\left[ 1+\ln f_M(v,t) \right]\dot f_M(v,t) \mathrm{d}^3v  \label{dH1},
\end{equation}{}
and applying the first order approximation (\ref{firstorder}) to (\ref{dH1}), we obtain
\begin{equation}
    \frac{dH'}{dt}=\sum_M\int_{}^{}\bar f(v) \left [ 1+\ln \left\{ \bar f(v)+\bar f(v)g_M(v,t) \right\} \right]\dot g_M(v,t)\mathrm{d}^3v  \label{dH1,1}.
\end{equation}{}
Using the same approximation (\ref{lnapproximation})
to (\ref{dH1,1}) 
\begin{eqnarray}
\frac{dH'}{dt}&=&\sum_M \int_{}^{} \bar f(v)\left[ 1+\ln \bar f(v)+g_M(v,t) \right]\dot g_M(v,t)\mathrm{d}^3v,
\end{eqnarray}{}
using (\ref{relacion1}) with homogeneous multipliers
\begin{eqnarray}
\frac{dH'}{dt}&=&\sum_M\int_{}^{}\bar f(v)\left[ \alpha+\beta \epsilon(v) \right]\dot g_M(v,t)\mathrm{d}^3v +\sum_M\int_{}^{}\bar f(v)g_M(v,t)\dot g_M(v,t)\mathrm{d}^3v \label{dH1,2}. \nonumber \\
\end{eqnarray}{}
If we use the same procedure shown in the previous section to the restrictions (\ref{restriccionescambio}), we obtain the same result. Then applying this to (\ref{dH1,2}), we find
\begin{eqnarray}
\frac{dH'}{dt}&=&\sum_M\int_{}^{}\bar f(v)g_M(v,t)\dot g_M(v,t)\mathrm{d}^3v.
\end{eqnarray}{}
Using the same consideration in the previous section to $g$ and $\dot g$, that is to divide the spatial integral in two, one integral for cells that have excess of particles and/or energy, and the second for cells that have missing of particles and/or energy, and then using (\ref{separacion}), we prove that $\frac{dH'}{dt}<0$. If the system is in equilibrium, $g(r,v,t)=\dot g(r,v,t)=0$, therefore $\frac{dH'}{dt}=0$. 
\\
\\
The next step in this model is to obtain the analogous transport equation through variation methods. In order to verify that transport equation can be obtained by applying the variation method, in the next section the Boltzmann transport equation will be obtained applying the mentioned method to the modified Boltzmann $H$ functional (\ref{CH2}).
